Relevance-based Quantization of Scattering Features for Unsupervised Mining of Environmental Audio

### 2018-02-02 paper/structureScene.tex
Edit scattering section

Scattering transforms are time-shift invariant representations of audio signals computed by applying auditory and modulation filter banks alternated with complex modulus nonlinearities.
This section first explains the importance of local invariance to time-shifting and stability to time-warping in the representation of acoustic scenes, then describes how the scattering transform is designed to satisfy these properties while having a high discriminative power.

\subsection{Invariance and stability in audio signals}
The notion of invariance to time-shifting plays an essential role in acoustic scene similarity retrieval as well as acoustic scene classification.
Indeed, recordings may be shifted locally without affecting their original class memberships.
To discard this superfluous source of variability, signals are mapped into a time-shifting invariant feature space before training the classifier, eliminating the need for this classifier to explicitly learn this invariance and increasing the robustness of its predictions.

@@

\subsection{Wavelet scalogram}
Our convention for the Fourier transform of a continuous-time signal $\boldsymbol{x}(t)$ is $\boldsymbol{\hat{x}}(\omega) = \int_{-\infty}^{+\infty} x(t) \exp(- \mathrm{i} 2\pi \omega t) \; \mathrm{d}t$.
Let $\boldsymbol{\psi}(t)$ a complex-valued band-pass filter of
center frequency $\xi_1$ and bandwidth $\xi_1/Q_1$, where $Q_1$ is the quality factor of the filter.
A filter bank of wavelets is built by dilating $\boldsymbol{\psi}(t)$
according to a geometric sequence of scales $2^{\gamma_1/Q_1}$, obtaining
\begin{equation}
\boldsymbol{\psi_{\gamma_1}}(t) = 2^{-\gamma_1/Q_1} \boldsymbol{\psi}(2^{-\gamma_1/Q_1} t)\mbox{.}
\end{equation}
The variable $\gamma_1$ is a scale (an inverse log-frequency) taking integer values between $0$ and $(J_1 \times Q_1 - 1)$, where $J_1$ is the number of octaves spanned by the filter bank.
For each $\gamma_1$, the wavelet $\boldsymbol{\psi_{\gamma_1}}(t)$
has a center frequency of $2^{-\gamma_1/Q_1}\xi_1$ and a bandwidth of $2^{-\gamma_1/Q_1}\xi_1/Q_1$ resulting in the same quality factor $Q_1$ as $\boldsymbol{\psi}$.
In the sequel, we set $\xi_1$ to $20~\mathrm{kHz}$, $J_1$ to $10$, and the quality factor $Q_1$, which is also the number of wavelets per octave, to $8$. This results in the wavelet filters covering the whole range of human hearing, from $20~\mathrm{Hz}$ to $20~\mathrm{kHz}$. Setting $Q_1 = 8$ results in filters whose bandwidth approximates an equivalent rectangular bandwidth (ERB) scale.

The wavelet transform of an audio signal
$\boldsymbol{x}(t)$ is obtained by convolution with all wavelet filters
Applying pointwise complex modulus the transform yields
the wavelet scalogram
\begin{equation}
\boldsymbol{x_1}(t, \gamma_1)
= \vert \boldsymbol{x} \ast \boldsymbol{\psi_{\gamma_1}} \vert (t)\mbox{.}
\end{equation}
The scalogram bears resemblance to the constant-Q transform (CQT),
which is derived from the short-term Fourier transform (STFT) by averaging the frequency
axis into constant-Q subbands of center frequencies $2^{-\gamma_1/Q_1}\xi_1$.
Indeed, both time-frequency representations are indexed by time $t$ and log-frequency $\gamma_1$.
However, contrary to the CQT, the scalogram reaches the Heisenberg
theoretical limit of optimal time-frequency localization across the whole
frequency range, whereas the temporal resolution of the traditional CQT is fixed by the support of the STFT analyzing window. %\cite{Brown1992}
Therefore, the scalogram has a better temporal localization at high
frequencies than the CQT, at the expense of a greater computational cost
since the inverse fast Fourier transform (IFFT) routine must be called for each wavelet $\boldsymbol{\psi_{\gamma_1}}$ in the filter bank.
However, this allows us to observe amplitude modulations at fine temporal scales in the scalogram, down to $2Q_1/\xi_1$ for $\gamma_1 = 0$, of the order of $1\,\textrm{ms}$ given the aforementioned values of $Q_1$ and $\xi_1$.

To obtain the desired invariance and stability properties, the scalogram is averaged in time using a lowpass filter $\phi(t)$ with cut-off frequency $1/T$ (and approximate duration $T$), to get
\begin{equation}
\mathbf{S_1}\boldsymbol{x}(t, \gamma_1) = \boldsymbol{x_1}(\cdot, \gamma_1) \ast \phi(t),
\end{equation}
which is known as the set of first-order scattering coefficients. They capture the average spectral envelope of $\boldsymbol{x}$ over scales of duration $T$, with spectral resolution varying with constant Q. In this way, they are closely related to the mel-frequency spectrogram and related features, such as MFCCs.

\subsection{Extracting modulations with second-order scattering}
In auditory scenes, short-time amplitude modulations may be caused by a variety of rapid mechanical interactions, including collision, friction, turbulent flow, and so on.
At longer time-scales, they also account for higher-level attributes of sound, such as prosody in speech or rhythm in music.
Although they are discarded while filtering $\boldsymbol{x_1}(t,\gamma_1)$ into a local time-shifting invariant representation $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$, they can be recovered by a second wavelet transform followed by a pointwise modulus operator.

We define second-order wavelets $\boldsymbol{\psi_{\gamma_2}}(t)$ in the same way as the first-order wavelets, but with parameters $\xi_2$, $J_2$, and $Q_2$ and so have center frequencies $2^{-\gamma_2/Q_2}\xi_2$ for $\gamma_2$ taking values between $0$ and $(J_2 \times Q_2 - 1)$. While this abuses notation slightly, the identity of the wavelets should be clear from context.
The amplitude modulation spectrum resulting from a wavelet modulus decomposition using these second-order wavelets is then
\begin{equation}
\boldsymbol{x_2}(t,\gamma_1,\gamma_2) =
\vert \boldsymbol{x_1} \ast \boldsymbol{\psi_{\gamma_2}} \vert(t,\gamma_1).
\end{equation}
In the sequel, we set $\xi_2$ to $2.5\,\mathrm{kHz}$, $Q_2$ to $1$, and $J_2$ to $12$. Lastly, the low-pass filter $\phi(t)$ is applied to $\boldsymbol{x_2}$ to guarantee local invariance to time-shifting, which yields
\begin{equation}
\mathbf{S_2}\boldsymbol{x}(t,\gamma_1,\gamma_2) =
(\boldsymbol{x_2}(\cdot,\gamma_1,\gamma_2) \ast \phi)(t).
\end{equation}

The scattering transform $\mathbf{S}\boldsymbol{x}(t,\gamma)$ consists of the concatenation of first-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$ and second-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1,\gamma_2)$ into a feature matrix $\mathbf{S}\boldsymbol{x}(t,\gamma)$, where $\gamma$ is a shorthand for either $\gamma_1$ or $(\gamma_1,\gamma_2)$. While higher-order scattering coefficients can be calculated, for the purposes of our current work, the first and second order are sufficient.

\subsection{Gammatone wavelets}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{gfx/gammatones}
\caption{
\label{fig:gammatones}
Gammatone wavelets $\psi(t)$ in the time domain with quality factors (a) $Q = 4$ and (b) $Q = 1$.
Oscillations represent the real and imaginary parts. The envelope represents the complex modulus.}
\end{center}
\end{figure}
Wavelets
$\boldsymbol{\psi_{\gamma_1}}(t)$ and $\boldsymbol{\psi_{\gamma_2}}(t)$ are designed as fourth-order Gammatone
wavelets with one vanishing moment \cite{Venkitaraman2014}, and are shown in Figure \ref{fig:gammatones}.
In the context of auditory scene analysis, the asymmetric envelope of Gammatone wavelets is more biologically plausible than the symmetric, Gaussian-like envelope of the more widely used Morlet wavelets.
Indeed, it allows to reproduce two important psychoacoustic effects in the mammalian cochlea: the asymmetry of temporal masking and the asymmetry of spectral masking  \cite{Fastl2007}.
The asymmetry of temporal masking is the fact that a masking noise has to be louder if placed after the onset of a stimulus rather than before.
Likewise, because critical bands are skewed towards higher frequencies, a masker tone has to be louder if it is above the stimulus in frequency rather than below.
It should also be noted that Gammatone wavelets follow the typical amplitude profile of natural sounds, beginning with a relatively sharp attack and ending with a slower decay.
As such, they are similar to filters discovered automatically by unsupervised encoding of natural sounds \cite{Smith2006}.
This suggests that, despite being hand-crafted and not learned, Gammatone wavelets provide a sparser time-frequency representation of acoustic scenes compared to other variants.

>>>>

Scattering transforms are time-shift invariant representations of audio signals computed by applying auditory and modulation wavelet filter banks alternated with complex modulus nonlinearities.
This section first explains the importance of local invariance to time-shifting and stability to time-warping in the representation of acoustic scenes, then describes how the scattering transform is designed to satisfy these properties while retaining high discriminative power.

\subsection{Invariance and stability in audio signals}
The notion of invariance to time-shifting plays an essential role in acoustic scene similarity retrieval as well as acoustic scene classification.
Indeed, recordings may be shifted locally without affecting their similarity to other recordings.
To discard this superfluous source of variability, signals are first mapped into a time-shifting invariant feature space. These features are then used to calculate similarities or to train a classifier. Since invariance is ensured by the features, it does not have to be learned during this last model construction step.

@@

\subsection{Wavelet scalogram}
Our convention for the Fourier transform of a continuous-time signal $\boldsymbol{x}(t)$ is $\boldsymbol{\hat{x}}(\omega) = \int_{-\infty}^{+\infty} x(t) \exp(- \mathrm{i} 2\pi \omega t) \; \mathrm{d}t$.
Let $\boldsymbol{\psi}(t)$ a complex-valued analytic bandpass filter of
center frequency $\xi_1$ and bandwidth $\xi_1/Q_1$, where $Q_1$ is the quality factor of the filter.
A filter bank of wavelets is built by dilating $\boldsymbol{\psi}(t)$
according to a geometric sequence of scales $2^{\gamma_1/Q_1}$, obtaining
\begin{equation}
\boldsymbol{\psi_{\gamma_1}}(t) = 2^{-\gamma_1/Q_1} \boldsymbol{\psi}(2^{-\gamma_1/Q_1} t)\mbox{.}
\end{equation}
The variable $\gamma_1$ is a scale (an inverse log-frequency) taking integer values between $0$ and $(J_1 Q_1 - 1)$, where $J_1$ is the number of octaves spanned by the filter bank.
For each $\gamma_1$, the wavelet $\boldsymbol{\psi_{\gamma_1}}(t)$
has a center frequency of $2^{-\gamma_1/Q_1}\xi_1$ and a bandwidth of $2^{-\gamma_1/Q_1}\xi_1/Q_1$ resulting in the same quality factor $Q_1$ as $\boldsymbol{\psi}$.
In the following, we set $\xi_1$ to $20~\mathrm{kHz}$, $J_1$ to $10$, and the quality factor $Q_1$, which is also the number of wavelets per octave, to $8$. This results in the wavelet filters covering the whole range of human hearing, from $20~\mathrm{Hz}$ to $20~\mathrm{kHz}$. Setting $Q_1 = 8$ results in filters whose bandwidth approximates an equivalent rectangular bandwidth (ERB) scale \cite{Fastl2007}.

The wavelet transform of an audio signal
$\boldsymbol{x}(t)$ is obtained by convolution with all wavelet filters.
Applying pointwise complex modulus the transform yields
the wavelet scalogram
\begin{equation}
\boldsymbol{x_1}(t, \gamma_1)
= \vert \boldsymbol{x} \ast \boldsymbol{\psi_{\gamma_1}} \vert (t)\mbox{.}
\end{equation}
The scalogram bears resemblance to the constant-Q transform (CQT),
which is derived from the short-term Fourier transform (STFT) by averaging the frequency
axis into constant-Q subbands of center frequencies $2^{-\gamma_1/Q_1}\xi_1$.
Indeed, both time-frequency representations are indexed by time $t$ and log-frequency $\gamma_1$.
However, contrary to the CQT, the scalogram reaches a better time-frequency localization across the whole
frequency range, whereas the temporal resolution of the traditional CQT is fixed by the support of the STFT analyzing window. %\cite{Brown1992}
Therefore, the scalogram has a better temporal localization at high
frequencies than the CQT, at the expense of a greater computational cost
since the inverse fast Fourier transform routine must be called for each wavelet $\boldsymbol{\psi_{\gamma_1}}$ in the filter bank.
However, this allows us to observe amplitude modulations at fine temporal scales in the scalogram, down to $2Q_1/\xi_1$ for $\gamma_1 = 0$, of the order of $1\,\textrm{ms}$ given the aforementioned values of $Q_1$ and $\xi_1$.

To obtain the desired invariance and stability properties, the scalogram is averaged in time using a lowpass filter $\phi(t)$ with cut-off frequency $1/T$ (and approximate duration $T$), to get
\begin{equation}
\mathbf{S_1}\boldsymbol{x}(t, \gamma_1) = \boldsymbol{x_1}(\cdot, \gamma_1) \ast \phi(t),
\end{equation}
which is known as the set of first-order scattering coefficients. They capture the average spectral envelope of $\boldsymbol{x}(t)$ over scales of duration $T$ and where the spectral resolution varying with constant Q. In this way, they are closely related to the mel-frequency spectrogram and related features, such as MFCCs.

\subsection{Extracting modulations with second-order scattering}
In auditory scenes, short-time amplitude modulations may be caused by a variety of rapid mechanical interactions, including collision, friction, turbulent flow, and so on.
At longer time-scales, they also account for higher-level attributes of sound, such as prosody in speech or rhythm in music.
Although they are discarded while filtering $\boldsymbol{x_1}(t,\gamma_1)$ into the time-shift invariant representation $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$, they can be recovered by a second wavelet transform and another complex modulus.

We define second-order wavelets $\boldsymbol{\psi_{\gamma_2}}(t)$ in the same way as the first-order wavelets, but with parameters $\xi_2$, $J_2$, and $Q_2$. Consequently, they have center frequencies $2^{-\gamma_2/Q_2}\xi_2$ for $\gamma_2$ taking values between $0$ and $(J_2 Q_2 - 1)$. While this abuses notation slightly, the identity of the wavelets should be clear from context.
The amplitude modulation spectrum resulting from a wavelet modulus decomposition using these second-order wavelets is then
\begin{equation}
\boldsymbol{x_2}(t,\gamma_1,\gamma_2) =
\vert \boldsymbol{x_1} \ast \boldsymbol{\psi_{\gamma_2}} \vert(t,\gamma_1).
\end{equation}
In the following, we set $\xi_2$ to $2.5\,\mathrm{kHz}$, $Q_2$ to $1$, and $J_2$ to $12$. Lastly, the low-pass filter $\phi(t)$ is applied to $\boldsymbol{x_2}(t, \gamma_1, \gamma_2)$ to guarantee local invariance to time-shifting, which yields the second-order scattering coefficients
\begin{equation}
\mathbf{S_2}\boldsymbol{x}(t,\gamma_1,\gamma_2) =
(\boldsymbol{x_2}(\cdot,\gamma_1,\gamma_2) \ast \phi)(t).
\end{equation}

The scattering transform $\mathbf{S}\boldsymbol{x}(t,\gamma)$ consists of the concatenation of first-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$ and second-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1,\gamma_2)$ into a feature matrix $\mathbf{S}\boldsymbol{x}(t,\gamma)$, where $\gamma$ is a shorthand for either $\gamma_1$ or $(\gamma_1,\gamma_2)$. While higher-order scattering coefficients can be calculated, for the purposes of our current work, the first and second order are sufficient.
Indeed, higher-order scattering coefficients have been shown to contain reduced energy and are therefore of limited use \cite{irene}.

\subsection{Gammatone wavelets}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{gfx/gammatones}
\caption{
\label{fig:gammatones}
Gammatone wavelets $\psi(t)$ in the time domain with quality factors (a) $Q = 4$ and (b) $Q = 1$.
Oscillations (red, blue) are the real and imaginary parts. The envelope (yellow) is the complex modulus.}
\end{center}
\end{figure}
Wavelets
$\boldsymbol{\psi_{\gamma_1}}(t)$ and $\boldsymbol{\psi_{\gamma_2}}(t)$ are designed as fourth-order Gammatone
wavelets with one vanishing moment \cite{Venkitaraman2014}, and are shown in Figure \ref{fig:gammatones}.
In the context of auditory scene analysis, the asymmetric envelopes of Gammatone wavelets are more biologically plausible than the symmetric, Gaussian envelopes of the more widely used Morlet wavelets.
Indeed, it allows to reproduce two important psychoacoustic effects in the mammalian cochlea: the asymmetry of temporal masking and the asymmetry of spectral masking  \cite{Fastl2007}.
The asymmetry of temporal masking is the fact that a masking noise has to be louder if placed after the onset of a stimulus rather than before.
Likewise, because critical bands are skewed towards higher frequencies, a masker tone has to be louder if it is above the stimulus in frequency rather than below.

It should also be noted that Gammatone wavelets follow the typical amplitude profile of natural sounds, beginning with a relatively sharp attack and ending with a slower decay.
As such, they are similar to filters discovered automatically by unsupervised encoding of natural sounds \cite{Smith2006}.
This suggests that, despite being hand-crafted and not learned, Gammatone wavelets provide a sparser time-frequency representation of acoustic scenes compared to other variants.
